SIMD-vectorized Chemical Source Term Evaluation

### 2017-02-25 paper.tex
larger plots
minor updates

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for states $1, 2\ldots$ will load sequential locations in memory, however the $(k+1)$th species concentration will be $N_{\text{state}}$ memory locations away increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state $j$, thus SIMD-lanes loading species concentrations $k, k+1\ldots$ will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU.
However, a deep vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).

@@

\subsection{Previous works and goals of this study}
Several recent works, e.g. \cite{CurtisGPU:2017}, have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code.

@@

In computing the net ROP of reaction $i$ for state $j$ from the forward and reverse ROP: $R_{ij} = R_{ij}^{\prime} - R_{ij}^{\prime\prime}$, precision can easily be lost as the net ROP may be---particularly near chemical equilibrium---many orders of magnitude smaller than the forward or reverse rates.
A base-ten bound on loss of precision~\cite{goldberg1991every} can be found as:
\begin{equation}
\text{L}_{ij,\text{lower}} = 2^{-p_i} \le \left\lvert 1 - \frac{R_{ij}^{\prime}}{R_{ij}^{\prime\prime}} \right\rvert \le 2^{-q_i} = \text{L}_{ij,\text{upper}}
\label{e:bounds}
\end{equation}
with $q_i \le p_i$.
To obtain an estimate of the precision bounds, the forward and reverse ROP from Cantera~\cite{Cantera} are substituted into Eq.~\eqref{e:bounds}.
Additionally a higher-precision number format (in this case, a quad-precision floating point) is used in this computation to maintain accuracy.
Finally, when determining the error in the net ROP between this code and Cantera~\cite{Cantera}, the bounds on the percent error attributable to precision loss, $\text{P}_{i,\text{err}}$ can be estimated as:
\begin{equation}
\text{P}_{ij,\text{lower}} = 100 * \frac{\text{L}_{ij,\text{lower}}}{\left\lvert\ R_{ij,\text{CT}} - R_{ij}\right\rvert} \le
\text{P}_{ij,\text{err}} \le 100 * \frac{\text{L}_{ij,\text{upper}}}{\left\lvert R_{ij,\text{CT}} - R_{ij} \right\rvert} = \text{P}_{ij,\text{upper}}
\label{e:rel_bound}
\end{equation}
where the \text{CT} subscript indicates values from Cantera~\cite{Cantera}.
In the case that $R_{ij,\text{CT}} = R_{ij}$, the error of this net ROP is zero and the precision loss is ignored.

For a more direct comparison, a relative error norm of a quantity $X_{ij}$ over all states $j$, and reactions $i$ was computed using the $L^{\infty}$ norm:
\begin{equation}
 E_{X} = \left\lVert \frac{\left\lvert X_{ij,\text{CT}} - X_{ij}\right\rvert}{\num{e-10} + \num{e-6} * \left\lvert X_{i,j,\text{CT}} \right\rvert} \right\rVert_{i,j,\infty}
\label{e:rel_err}
\end{equation}
For the reaction $i$ and state $j$ that results in the maximum ROP error $E_{R}$, the lower precision bound is set to $\text{P}_{\text{lower}}^{*} = \text{P}_{i,\text{lower}}$.

@@

In Table~\ref{T:error}, we see the results of this code as compared to Cantera~\cite{Cantera} on a library of partially stirred reaction conditions (PaSR) described in our previous works~\cite{CurtisGPU:2017,Niemeyer:2016aa}.
Very close agreement is observed for the forward and reverse ROP for all models, however the net ROP error norm is \numrange{6}{8} orders of magnitude larger.
We see here that the estimated lower bound on loss of precision is \SIrange{4}{40}{$\times$} the error in net ROP, explaining the large increase in the net ROP error norm as compared to either the forward or backward ROP error norms.
The species rate error norm is similar in magnitude to that of the net ROP, however the temperature rate error again increases as it depends directly on the species rates.
We note that the above discussion does not imply that computation of the net ROP will cause large error in chemical kinetic integration---either in this code or Cantera~\cite{Cantera}---as this loss of precision only occurs when the forward and reverse ROP are nearly equal, implying the reaction is in near-equilibrium.

@@

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_SIMT_Order_comparison}
    \caption{Runtime in milliseconds per thermo-chemical state of shallow-SIMD and SIMT-vectorization on a single CPU core for both ``C'' (filled symbols) and ``F'' (empty symbols) data ordering.}
    \label{F:SIMDComp}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_Vecwidth_comparison}
    \caption{Speedup of a ``C''-ordered shallow-SIMD vectorization on a single CPU core when varying the supplied vector-width.}
    \label{F:Veccomp}
  \end{subfigure} \\
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{SIMD_scaling}
    \caption{Strong parallel scaling of a ``C''-ordered shallow-SIMD vectorization using a vector-width of 16.}
    \label{F:SIMDscale}
  \end{subfigure}
\end{figure}

>>>>

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for states $1, 2\ldots$ will load sequential locations in memory, however the $(k+1)$th species concentration will be $N_{\text{state}}$ memory locations away increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state $j$, thus SIMD-lanes loading species concentrations $k, k+1\ldots$ for e.g., state 1, will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU.
However, a deep vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).

@@

\subsection{Previous works and goals of this study}
Several recent works, e.g.~\cite{CurtisGPU:2017}, have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including CPUs, GPUs, and MICs.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code.

@@

In computing the net ROP of reaction $i$ for state $j$ from the forward and reverse ROP: $R_{ij} = R_{ij}^{\prime} - R_{ij}^{\prime\prime}$, precision can easily be lost as the net ROP may be---particularly near chemical equilibrium---many orders of magnitude smaller than the forward or reverse rates.
A base-ten bound on loss of precision~\cite{goldberg1991every} can be found as:
\begin{equation}
\text{L}_{ij,\text{lower}} = 2^{-p_{ij}} \le \left\lvert 1 - \frac{R_{ij}^{\prime}}{R_{ij}^{\prime\prime}} \right\rvert \le 2^{-q_{ij}} = \text{L}_{ij,\text{upper}}
\label{e:bounds}
\end{equation}
with $q_{ij} \le p_{ij}$.
To obtain an estimate of the precision bounds, the forward and reverse ROP from Cantera~\cite{Cantera} are substituted into Eq.~\eqref{e:bounds}.
Additionally a higher-precision number format (in this case, a quad-precision floating point) is used in this computation to maintain accuracy.
Finally, when determining the error in the net ROP between this code and Cantera~\cite{Cantera}, the bounds on the percent error attributable to precision loss, $\text{P}_{ij,\text{err}}$ can be estimated as:
\begin{equation}
\text{P}_{ij,\text{lower}} = 100 * \frac{\text{L}_{ij,\text{lower}}}{\left\lvert\ R_{ij,\text{CT}} - R_{ij}\right\rvert} \le
\text{P}_{ij,\text{err}} \le 100 * \frac{\text{L}_{ij,\text{upper}}}{\left\lvert R_{ij,\text{CT}} - R_{ij} \right\rvert} = \text{P}_{ij,\text{upper}}
\label{e:rel_bound}
\end{equation}
where the \text{CT} subscript indicates values from Cantera~\cite{Cantera}.
In the case that $R_{ij,\text{CT}} = R_{ij}$, the error of this net ROP is zero and the precision loss is ignored.

For a more direct comparison, a relative error norm of a quantity $X_{ij}$ over all states $j$, and reactions $i$ was computed using the $L^{\infty}$ norm:
\begin{equation}
 E_{X} = \left\lVert \frac{\left\lvert X_{ij,\text{CT}} - X_{ij}\right\rvert}{\num{e-10} + \num{e-6} * \left\lvert X_{ij,\text{CT}} \right\rvert} \right\rVert_{i,j,\infty}
\label{e:rel_err}
\end{equation}
For the reaction $i$ and state $j$ that results in the maximum ROP error $E_{R}$, the lower precision bound is set to $\text{P}_{\text{lower}}^{*} = \text{P}_{i,\text{lower}}$.

@@

In Table~\ref{T:error}, we see the results of this code as compared to Cantera~\cite{Cantera} on a library of partially stirred reaction conditions (PaSR) described in our previous works~\cite{CurtisGPU:2017,Niemeyer:2016aa}.
Very close agreement is observed for the forward and reverse ROP for all models, however the net ROP error norm is \numrange{6}{8} orders of magnitude larger.
We see here that the estimated lower bound on loss of precision is \SIrange{4}{40}{$\times$} the error in net ROP, explaining the large increase in the net ROP error norm as compared to either the forward or backward ROP error norms.
The species rate error norm is similar in magnitude to that of the net ROP, however the temperature rate error again increases as it depends directly on the species rates.
We note that the above discussion does not imply that computation of the net ROP will cause large error in chemical kinetic integration---either in this code or Cantera~\cite{Cantera}---as this loss of precision only occurs when the forward and reverse ROP are nearly equal, implying the reaction is in near-equilibrium.
\todo{update for new valid}

@@

\begin{figure}[htb]
  \centering
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_SIMT_Order_comparison}
    \caption{Runtime in milliseconds per thermo-chemical state of shallow-SIMD and SIMT-vectorization on a single CPU core for both ``C'' (filled symbols) and ``F'' (empty symbols) data ordering.}
    \label{F:SIMDComp}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{SIMD_Vecwidth_comparison}
    \caption{Speedup of a ``C''-ordered shallow-SIMD vectorization on a single CPU core when varying the supplied vector-width.}
    \label{F:Veccomp}
  \end{subfigure} \\
  \begin{subfigure}[t]{0.5\textwidth}
    \includegraphics[width=\textwidth]{SIMD_scaling}
    \caption{Strong parallel scaling efficiency of a ``C''-ordered shallow-SIMD vectorization using a vector-width of 16.}
    \label{F:SIMDscale}
  \end{subfigure}
\end{figure}
